format

### 2021-05-04 tidymodels_analysis.Rmd
updated analysis - variable correlations and one-way ANOVAs

library(tidyverse)
library(nortest)
library(FSA)
library(tidymodels)
library(factoextra)
library(plotly)
library(usemodels)
library(ranger)
library(vip)

@@

## Dysregulation of hemoglobin and oxygenation in cerebral malaria {.tabset}

@@

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  select(-c(Years, Months))

@@

### Exploratory analysis

@@

### PCA

@@

# convert tibble to df
df_hbox_impute_df <- df_hbox_impute %>% select(-c(subject_id, Status)) %>% as.data.frame(df_hbox_impute) 
rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id

@@

p_pca <- fviz_pca_ind(my_pca,
                      col.ind = groups, # color by groups
                      palette = c("#00AFBB",  "#FC4E07", "#7CAE00"),
                      addEllipses = TRUE, # Concentration ellipses
                      ellipse.type = "confidence",
                      legend.title = "Groups",
                      repel = TRUE
)

p_pca$layers[[4]] <- NULL # get rid of text labels

p_pca

@@

# Biplot of individuals and variables
p_biplot <- fviz_pca_biplot(my_pca, repel = TRUE,
                            col.var = "black", # Variables color
                            col.ind = groups  # Individuals color
)

p_biplot$layers[[3]] <- NULL
p_biplot$layers[[2]] <- NULL

p_biplot

@@

df_pca <- pca_res_ind$coord %>% 
  as_tibble(rownames = "subject") %>% 
  mutate(Status = df_hbox_impute$Status)

@@

### Random forest predictive modeling: all data

@@

# Data
hbox_df <- df_hbox_trim %>% select(-subject_id)

@@

# computed on test set
collect_metrics(hbox_fit) %>% select(.metric, .estimate) %>% as.data.frame()

# collect predictions on test set
df_pred <- collect_predictions(hbox_fit) %>%
  select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))

df_pred %>% as_data_frame()

@@

  ggplot(aes(x = Truth, y = Freq, fill = Prediction)) + 
  geom_col(position = "dodge") +
  ggtitle("Confusion matrix") +
  theme_bw()

@@

### Random forest predictive modeling: NIRS data only

@@

# Data
hbox_df2 <- df_hbox_impute %>% 
  select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))

@@

# computed on test set
collect_metrics(final_rs_res2) %>% select(.metric, mean, std_err, n) %>% as.data.frame()

# collect predictions on test set
df_pred2 <- collect_predictions(hbox_fit2) %>%
  select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))

df_pred2 %>% as_data_frame()

@@

  ggplot(aes(x = Truth, y = Freq, fill = Prediction)) + 
  geom_col(position = "dodge") +
  ggtitle("Confusion matrix") +
  theme_bw()

@@

>>>>

library(tidyverse)
library(nortest)
library(FSA)
library(tidymodels)
library(factoextra)
library(ggpubr)
library(ggrepel)
library(grid)
library(pheatmap)
library(plotly)
library(corrr)
library(reshape)
library(RColorBrewer)
library(usemodels)
library(ranger)
library(vip)
library(MASS)

@@

# Dysregulation of hemoglobin and oxygenation in cerebral malaria {.tabset}

@@

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  dplyr::select(-c(Years, Months))

@@

## Exploratory analysis

@@

## PCA

@@

# convert tibble to df
df_hbox_impute_df <- df_hbox_impute %>% dplyr::select(-c(subject_id, Status)) %>% as.data.frame(df_hbox_impute) 
rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id

@@

df_pnts <- my_pca$x %>% as_tibble() %>% mutate(group = groups)

df_pnts %>%

  ggplot(aes(x = PC1, y = PC2, color = group, fill = group)) +
  geom_point(size = 3) +
  stat_conf_ellipse(level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(aes(shape = group), na.rm = FALSE) +
  geom_vline(xintercept = 0, lty = 2, color = "black") +
  geom_hline(yintercept = 0, lty = 2, color = "black") +

  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC1: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +

  theme_bw()

@@

df_vars <- my_pca$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca$rotation))

# plot

  # plot patients
  ggplot() +
  geom_point(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group), 
             size = 3) +
  stat_conf_ellipse(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +

  # plot variables
  geom_point(data = df_vars, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

  # draw lines
  geom_vline(xintercept = 0, lty = 2, color = "black") +
  geom_hline(yintercept = 0, lty = 2, color = "black") +

  # design
  xlab(paste0("PC1: ", round(pca_eig_val$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC1: ", round(pca_eig_val$variance.percent[2], 2), "% of variance")) +

  theme_bw()

@@

df_pca <- pca_res_ind$coord %>% 
  as_tibble(rownames = "subject") %>% 
  mutate(Status = factor(df_hbox_impute$Status, levels = c("UM", "HV", "CM")))

@@

## Variable correlations

```{r correlation, fig.heigh = 10, fig.width = 10}

m_cor <- df_hbox_impute %>% dplyr::select(-c(subject_id, Status)) %>% cor()

melt(m_cor) %>% 
  ggplot(aes(x = X1, y = X2, fill = value, label = round(value,2))) + 
  geom_tile() +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
  geom_text(size = 3) +

  xlab("") +
  ylab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.9))

## Random forest predictive modeling {.tabset}

### All variables

@@

# Data
hbox_df <- df_hbox_trim %>% dplyr::select(-subject_id)

@@

# computed on test set
collect_metrics(hbox_fit) %>% dplyr::select(.metric, .estimate) %>% as.data.frame()

# collect predictions on test set
df_pred <- collect_predictions(hbox_fit) %>%
  dplyr::select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))

df_pred %>% as.data.frame()

@@

  ggplot(aes(x = Truth, y = Freq, fill = Prediction, label = Freq)) + 
  geom_col(position = "dodge") +
  geom_text(position = position_dodge(width = 0.9)) +
  ggtitle("Confusion matrix") +
  theme_bw()

@@

### NIRS variables only

@@

# Data
hbox_df2 <- df_hbox_impute %>% 
  dplyr::select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))

@@

# computed on test set
collect_metrics(final_rs_res2) %>% dplyr::select(.metric, mean, std_err, n) %>% as.data.frame()

# collect predictions on test set
df_pred2 <- collect_predictions(hbox_fit2) %>%
  dplyr::select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))

df_pred2 %>% as.data.frame()

@@

  ggplot(aes(x = Truth, y = Freq, fill = Prediction, label = Freq)) + 
  geom_col(position = "dodge") +
  geom_text(position = position_dodge(width = 1)) +
  ggtitle("Confusion matrix") +
  theme_bw()

@@

### Other variables

#### Build, tune, and finalize model
Same code, but built model only using other variables (Arginine, lactate, hematocrit, temperature, whole blood nitrite, haptoglobin, glucose)

```{r model3}

# Data
hbox_df3 <- df_hbox_impute %>% dplyr::select(-c(subject_id, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd))

  # split data set
  set.seed(1234)
  hbox_split3 <- initial_split(hbox_df3, strata = Status)
  hbox_train3 <- training(hbox_split3)
  hbox_test3 <- testing(hbox_split3)

  # resamples (not a lot of data, need to use cross validation)
  set.seed(2345)
  hbox_folds3 <- bootstraps(hbox_train3, strata = Status)

  # Create recipe
  ranger_recipe3 <- 
    recipe(formula = Status ~ ., data = hbox_df3) %>%

  # Model specifications (set for tuning)
  ranger_spec3 <- 
    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
    set_mode("classification") %>% 
    set_engine("ranger")

  # Create workflow
  ranger_workflow3 <- 
    workflow() %>% 
    add_recipe(ranger_recipe3) %>% 
    add_model(ranger_spec3)

  # Initial tune on bootstraps
  set.seed(8434)
  ranger_tune3 <-
    tune_grid(ranger_workflow3, 
              resamples = hbox_folds3, 
              grid = 11)

  # finalize workflow
  final_rf3 <- ranger_workflow3 %>% 
    finalize_workflow(select_best(ranger_tune3, metric = "accuracy"))

  final_rs_res3 <- final_rf3 %>% 
    fit_resamples(resamples = hbox_folds3,
                  metrics = metric_set(roc_auc, accuracy, sens, spec),
                  control = control_resamples(save_pred = TRUE))

  hbox_fit3 <- last_fit(final_rf3, hbox_split3) # fitting to training data, evaluating on testing

```{r model3_res}

# computed on test set
collect_metrics(final_rs_res3) %>% dplyr::select(.metric, mean, std_err, n) %>% as.data.frame()

# collect predictions on test set
df_pred3 <- collect_predictions(hbox_fit3) %>%
  dplyr::select(c(Status, .pred_HV, .pred_UM, .pred_CM, .pred_class))

df_pred3 %>% as.data.frame()

df_pred3 %>% filter(Status != .pred_class) %>% as.data.frame()

```{r model3_plot_res}

conf_mat_resampled(final_rs_res3) %>%

  ggplot(aes(x = Truth, y = Freq, fill = Prediction, label = Freq)) + 
  geom_col(position = "dodge") +
  geom_text(position = position_dodge(width = 1)) +
  ggtitle("Confusion matrix") +
  theme_bw()

#### Determine variable importance
```{r model3_imp}

# new specification for ranger model
imp_spec3 <- ranger_spec3 %>% 
  finalize_model(select_best(ranger_tune3, metric = "accuracy")) %>% 
  set_engine("ranger", importance = "permutation")

# plot importance 
workflow() %>% 
  add_recipe(ranger_recipe3) %>% 
  add_model(imp_spec3) %>% 
  fit(hbox_train3) %>% 
  pull_workflow_fit() %>% 
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +

## One way ANOVAs {.tabset}

### Influence healthy vs malaria

#### Arginine
```{r arginine_show, echo = TRUE, eval = FALSE}

# create model for BoxCox transformation
arg_model <- glm(`Arginine (umol/L)` ~ Status, data = df_hbox_impute)
boxcox(arg_model, lambda = seq(-0.5, 1, by = 0.1), plotit = TRUE)

# BoxCox transform
df_arg <- df_hbox_impute %>% mutate(arg_transform = (`Arginine (umol/L)`^-0.25 - 1)/-0.25)

# model on transformed data
glm(arg_transform ~ Status, data = df_arg) %>% summary()

```{r arg_model}

glm(`Arginine (umol/L)` ~ Status, data = df_hbox_impute) %>% summary()

aov(`Arginine (umol/L)` ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r arg_plot}

df_hbox_long %>% filter(measurement == "Arginine (umol/L)") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("Arginine (umol/L)") +

#### Temperature
```{r temp}

glm(Temperature ~ Status, data = df_hbox_impute) %>% summary()

aov(Temperature ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r temp_plot}

df_hbox_long %>% filter(measurement == "Temperature") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("Temperature") +

#### Hematocrit
```{r hct}

glm(Hematocrit ~ Status, data = df_hbox_impute) %>% summary()

aov(Temperature ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r hct_plot}

df_hbox_long %>% filter(measurement == "Hematocrit") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("Hematocrit") +

### Influence CM vs UM

#### O2sat_alpha
```{r O2sat_alpha}

glm(O2sat_alpha ~ Status, data = df_hbox_impute) %>% summary()

aov(O2sat_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r O2sat_alpha_plot}

df_hbox_long %>% filter(measurement == "O2sat_alpha") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("O2sat_alpha") +

#### THC_alpha
```{r THC_alpha}

glm(THC_alpha ~ Status, data = df_hbox_impute) %>% summary()

aov(THC_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r THC_alpha_plot}

df_hbox_long %>% filter(measurement == "THC_alpha") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("THC_alpha") +

#### THC
```{r THC}

glm(THC ~ Status, data = df_hbox_impute) %>% summary()

aov(THC ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r THC_plot}

df_hbox_long %>% filter(measurement == "THC") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("THC") +

#### OxyHb
```{r OxyHb}

glm(OxyHb ~ Status, data = df_hbox_impute) %>% summary()

aov(OxyHb ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r OxyHb_plot}

df_hbox_long %>% filter(measurement == "OxyHb") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("OxyHb") +

#### Lactate
```{r Lactate}

glm(Lactate ~ Status, data = df_hbox_impute) %>% summary()

aov(Lactate ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r Lactate_plot}

df_hbox_long %>% filter(measurement == "Lactate") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("Lactate") +

#### ?OxyHb_alpha
```{r OxyHb_alpha}

glm(OxyHb_alpha ~ Status, data = df_hbox_impute) %>% summary()

aov(OxyHb_alpha ~ Status, data = df_hbox_impute) %>% TukeyHSD() %>% 
  .$Status %>% 
  as.data.frame() %>% 
  rownames_to_column("Status") %>% 
  dplyr::select(Status, `p adj`)

```{r OxyHb_alpha_plot}

df_hbox_long %>% filter(measurement == "OxyHb_alpha") %>%

  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +

  theme_bw() +
  theme(legend.position = "none")  +
  ggtitle("OxyHb_alpha") +
